#' ---
title: "Overview of LGT&SGT Model"
author: "Wibowo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of LGT&SGT Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Overview of LGT&SGT models
This vignette will provide an overview of the theoretical construction of LGT and SGT family models. A more complete description will be provided in the accompanying research paper which is planned to be released in 2019.The idea of the construction of LGT & SGT models originally came from Smyl in 2016 and has evolved over time in the collaboration process of writing this package.

The "LGT" model stands for Local-Global Trend and is devised to forecast on non-seasonal time series data. Meanwhile, the "SGT" model stands for Seasonal-Global Trend and is devised to complement the former model on seasonal time series data. The combination of these models were tested on the 3003 time-series data, available from the M3-competition dataset, and has been shown to outperform all of the models originally participating in the competition by a significant margin.
 
| Model         | Non-seasonal  | Seasonal      | Aggregate     |
|---------------|sMAPE  |MASE   |sMAPE  |MASE   |sMAPE  |MASE   |
|===============|===============|===============|===============|
|Dampen-M3      |14.6508|2.4689 |12.7710|2.0686 |12.3827|2.1778 |
|Best M3 Model  |       |       |       |       |12.76  |1.39   |
|Dampen-fpp2    |14.3203|2.3702 |11.6795|1.3588 |12.3997|1.6346 |
|LGT&SGT        |12.6732|2.0280 |12.0450|1.1265 |12.2163|1.3724 |
*note: The details of the benchmark models are as follow:
1. Dampen-M3: The implementation of Dampen algtortihm as it appears in original M3 competition.
2. Best M3 model: In this case, the best model is Theta (cite paper)
3. Dampen-fpp2 : An implementation of ETS model from fpp2, a popular forecasting package available from CRAN (cite Hyndmann)
The explanations of sMAPE and MASE can be found in ....

The LGT&SGT models are inherently Bayesian ETS (Exponential Smoothing) models which are implemented with the aid of RStan package. As bayesian models, the LGT&SGT models will require specification of prior distributions (with default priors provided) and might also take extended running time in some unusual cases. Generally, the running time should be acceptable if the models are run with the default prior and sampling algorithm. In unusual circumstances, some modifications on the prior distributions and Bayesian sampling algorithm might be required to speed up the running time of the models.

As a caveat, the current version of the models only support time series data with entirely positive data values. Many time series data fall into this category, while for others, simple transformations can be performed to obtain the required positive time-series data. Therefore, the constraint should not be too restrictive for most general applications.
 
 The following sections will briefly describe the mathematical constructions 
 and the rationale for these models.
 
## LGT(Local and Global Trend)

### subsection{Model Equations}{
In terms of mathematical notation, the model can be fully represented as follow:
 
$$y_{t+1} \sim Student (\nu,y_{t+1}, \sigma _{t+1}) \quad (eq.1.1)$$
$$\widehat{y}_{t+1}=l_{t}+ \gamma l_{t}^{ \rho }+ \lambda b_{t}  \quad  (eq. 1.2)$$
$$l_{t+1}= \alpha y_{t+1}+ \left( 1- \alpha  \right)  \left(  l_{t} \right) \quad (eq. 1.3)$$
$$b_{t+1}= \beta  \left( l_{t+1}-l_{t} \right) + \left( 1- \beta  \right) b_{t}  \quad  (eq. 1.4)$$
$$\widehat{\sigma}_{t+1}= \sigma l_{t}^{ \tau}+ \xi   \quad  (eq. 1.5)$$

### Notations
*$y_{t}$  :value of the dependent variable of interest at time t
*$\widehat{y}_{t+1}$  :forecasted value of y at time t+1 given information up to time t
*$\widehat{\sigma}_{t+1}$  :forecasted deviation at time t+1 given information up to time t
*$l_{t}$  :level at time t
*$b_{t}$  :local trend at time t

 
 
### Parameters
*$\nu$  :degrees of freedom of the t-distribution
*$\gamma$  :coefficient of the global trend
*$\rho$  :power coefficient of the global trend
*$\lambda$  :damping coefficient of the local trend
*$\alpha$  :smoothing parameter for the level term
*$\beta$  :smoothing parameter for the local trend term
*$\sigma$  :coefficient of the heteroscedastic standard deviation
*$\tau$  :power coefficient of the heteroscedastic standard deviation
*$\xi$  :base or minimum value of the standard deviation

### Rationale for Mathematical Equations
 
The LGT model is devised as an extension to the classical ETS models by incorporating the following additional features:

1. Heteroscedastic and non-normal error term
2. Addition of global trend term

The inclusion of these features on each individual equation of the model is discussed below.

#### Eq. 1.1. Student's-t Error Distribution}
 
The data value follows Student's t-distribution around the 
expected value of the data with a time-varying standard deviation. 
The Student-t distribution can be seen as a generalisation of the normal 
distribution to allow for a fat-tailed error distribution.
 
 
#### Eq. 1.5. Heteroscedastic Error
 
In addition to accounting for possible fat-tailed error distribution, 
the error function also allows the variance of the error to change as the level changes. This is achieved by allowing the scale (deviation) parameter of the assumed Student's t-distribution to vary in proportion to the current level of the time series. This will account for common situations, where the magnitude of the error will increase as the value of the data points increases.Moreover, this relationship does not necessarily be linear as the parameter $\tau$  controls the growth of the variance of the error term. 


In practice, the values taken by the parameter is often limited between 0 and 1. A value of 0 corresponds to constant variance, i.e. homoscedasticity, whereas a value of 1 approximates the behaviour of the multiplicative error ETS model. The value between 0 and 1 will describe an error function which grows in relation to the increase in data value, but at a slower pace than linear growth.
 
 
#### Eq. 1.2: One-step Ahead Prediction Equation}

There are three distinct terms that constitute this Bayesian ETS model: a level term, and a couple of different trends. The first term $l_{t}$ is the level term, while the second term  $l_{t}^{ \rho }$  refers to the global trend which increases with the level of the dependent variables  $l_{t}$  at a constant rate $\gamma$ . Similar to the heteroscedasticity in eq. 1.5, the change in the value of the dependant variable does not necessarily grow linearly with respect to the level of the variable. This relationship can be tuned in by the use of exponential parameter $\rho$. The interpretation of this global trend is also analogous to the time-varying error term, i.e. 
the value of  $ 0< \rho <1$  indicates a global trend which grows faster than the additive models but slower than the multiplicative model. The last term of the right-hand side of eq.1.2. refers to the usual local dampen trend in ETS model. However, there is an addition of dampen parameter $\lambda$, constrained such that  $-1<\lambda<1$, to reduce the strength of the local trend model.


#### Eq. 1.3: Level Adjustment Equation
 
This equation is defined according to the classical linear trend ETS model. 
The level at time t, $l_{t}$, is calculated as a weighted average of 
the current observation $y_{t}$  and the previous level at lag 1, $l_{t-1}$
with smoothing parameter $\alpha$.

#### Eq. 1.4: Local Trend Adjustment Equation

Similarly, the evolution of the local trend  $b_{t}$ is identically defined to the linear trend method. The local trend at time t  $b_{t}$  is obtained as a weighted average of the difference in level terms  $(l_{t}-l_{t-1})$  
and the trend at time t-1  $b_{t-1}$ with the smoothing parameter  $\beta$.


 
## SGT (Seasonal, Global Trend)
 
SGT model was designed as a seasonal counterpart to the LGT model. 
Similar to LGT, this model is devised to allow for global trend term and heteroscedastic error.
 
### Model Equations
 
$$y_{t+1} \sim Student \left( \nu,\widehat{y}_{t+1}, \sigma _{t+1} \right)  \quad (eq. 2.1) $$
$$\widehat{y}_{t+1}= \left( l_{t}+ \gamma l_{t}^{ \rho } \right)  s_{t+1} \quad (eq. 2.2)$$
$$l_{t+1}= \alpha  \frac{y_{t+1}}{s_{t+1}}+ \left( 1- \alpha  \right)  \left( l_{t} \right) \quad (eq. 2.3)$$ 
$$s_{t+m+1}= \zeta  \frac{y_{t+1}}{l_{t+1}}+ \left( 1- \zeta  \right) s_{t+1}  \quad (eq. 2.4)$$
$$\widehat{\sigma}_{t+1}= \sigma \widehat{y}_{t+1}^{ \tau}+ \xi \quad (eq. 2.5)$$
 
### Additional Notations

*$s_{t}$  : seasonality factor at time t
*$m$      : number of seasons in the data (e.g. 12 for monthly, 4 for quarterly)

### Additional Parameters
 
 \describe{
  \item{$\zeta}}{smoothing parameter for the seasonality terms}
 }
 
 }
 
 \subsection{Rationale for Mathematical Equations}{
 
 Most of the components of the model is similar to the non-seasonal model above. 
 A couple of modifications have been applied to this model in comparison to LGT:
 
 \enumerate{
 \item Removal of local dampen trend
 \item Addition of multiplicative seasonality term
 }
 
 The mathematical equations have also been adjusted accordingly:
 
 \itemize{
 \item \strong{Eq. 2.2.One-step Ahead Prediction Equation}
 
 Comparing eq. 2.2 with eq. 1.2, it is apparent that the local trend term  
 $\lambda b_{t}}  has been removed from the SGT model. 
 Consequently, the equation governing the evolution of local trend over time (eq. 2.4) is 
 no longer needed. Based on empirical evidence, the role of the local trend was 
 found to be insignificant for seasonal data in M3-competition dataset.
 
 Moreover, there is also an addition of multiplicative seasonality term modelled 
 after Holt-Winters seasonality term. In this case, the data value is decomposed 
 into level value and seasonal value (i.e.  { y_{t}=l_{t} s_{t} } ). Thus, it follows 
 that the one-step prediction value of  { y_{t+1} }  (eq. 2.2) is obtained by 
 multiplying the predicted level term  $l_{t}+ \gamma l_{t}^{ \rho }} with the 
 seasonal term  $s_{t+1}}. 

 \item \strong{Eq. 2.4.Seasonal Factors Adjustment Equation}
 
 The evolution of the seasonal component in eq. 2.4. is also based on standard 
 ETS model, i.e. a weighted average of predicted seasonal level  
 $\left( \frac{y_{t}}{l_{t}} \right)}  and previous seasonal 
 value from the past observations $s_{t}} .
 
 \item \strong{Eq. 2.3.Level Adjustment Equation}
 
 The evolution of level term in eq. 2.3 is defined analogously to the level 
 term in LGT model. The level term of time t  $\left( l_{t} \right)}  is 
 calculated as a weighted average of predicted current level value at time t  
 $ \frac{y_{t}}{s_{t}} }  and the previous estimate of the level at time t 
 given information up to t-1, i.e.  $ l_{t-1}+ \gamma l_{t-1}^{ \rho }.}  
 
  \item \strong{Eq. 2.5.Heteroscedastic Error}
  
 There is also a straightforward modification to the error term in SGT eq. 2.5 
 due to the introduction of seasonal factors. Compared to eq. 1.5, the error term
  in SGT is allowed to vary in proportion to $  y  } instead of  $ l_{t} } . 
  The reason for that is because the error term is likely to become larger d
  uring the peak seasons (i.e. seasons with high seasonality coefficient). 
  Hence, it is intuitive to link the dispersion of the error to the predicted 
  value of the dependent variable. Note that in the case of LGT, the predicted 
  value of y is also a function of the previous level  $ l_{t-1} }  and 
  thus, the error term can be defined in terms of  $ l_{t-1} }  instead.
 }
 }
 
 @section S2GT (Double Seasonal, Global Trend):
 
 S2GT is designed as an extension to SGT in time-series data which exhibit two seasonality patterns. 
 The additional second seasonality factor generalises the model to capture a number of seasonalities 
 that exist in the data. A classic example of this type of data is the hourly electricity consumption data 
 which are affected by the time of the day as well as the day in the week. 
 The mathematical modelling is based on the classical Double Seasonal Exponential Smoothing 
 model by Taylor (2003).
 
 \subsection{Model Equations}{
 
 \deqn{y_{t+1} \sim Student \left( v,y_{t+1}, \sigma _{t+1} \right) \quad (eq. 3.1)} 
 \deqn{ y_{t+1}=\left( l_{t}+ \gamma l_{t}^{ \rho } \right) s_{t+1}w_{t+1} \quad (eq. 3.2)} 
 \deqn{ l_{t}= \alpha  \frac{y_{t}}{s_{t}w_{t}}+ \left( 1- \alpha  \right) \left( l_{t-1} \right)  \quad (eq. 3.3)} 
 \deqn{ s_{t+m}= \zeta  \frac{y_{t}}{l_{t}w_{t}}+ \left( 1- \zeta  \right) s_{t} \quad (eq. 3.4)} 
 \deqn{ w_{t+d}= \delta  \frac{y_{t}}{l_{t}s_{t}}+ \left( 1- \delta  \right) w_{t} \quad (eq. 3.5)} 
 \deqn{  \sigma _{t+1}= \sigma y_{t+1}^{ \tau}+ \varsigma  \quad (eq. 3.6)} 
 
 }
 \subsection{Additional Notations}{
 \describe{
 \item{$ w_{t} }}{second seasonality factor prevailing at time t}
 \item{$ d }}{number of (second) seasons in a complete period (e.g. 12 for monthly, 4 for quarterly}
 }
 }
 
 \subsection{Additional Parameters}{
 \describe{
 \item{$\delta}}{smoothing parameters for the second seasonality factors}
 }
 }
 
 \subsection{Rationale for Mathematical Equations}{
 There is a minor difference in this model compared to the SGT model:
 
 \enumerate{
 \item Addition of second seasonality factors
 }
 
 The mathematical equations have been adjusted as follow:
 
 \itemize{
 
 \item \strong{Eq. 3.2. One-step Ahead Prediction Equation}

 Comparing eq. 2.2 with eq. 2.2, there is an addition of the second seasonality factor in multiplicative form. 
 Therefore, both seasonality factors will affect the proportionality between the expected 
 value of the observed data and the level. 
 
 \item \strong{Equation 3.3. Level Adjustment Equation}
 
 The evolution of level term in eq. 3.3 is largely identical to eq.2.3. 
 The only difference is that since the observed value is now seen as a product 
 of the level terms with both seasonality factors, the predicted current level 
 value at time t is now calculated as   $ \frac{y_{t}}{l_{t}s_{t}} }. 
 The weighted average in the right-hand side of the equation is then adjusted accordingly.
 
 \item \strong{Eq. 3.4. First Seasonal Factors Adjustment Equation}
 
 The evolution of the seasonal component in eq. 3.4. is also similar to eq. 2.4. 
 Again, the predicted seasonal level based on current observation has been changed to  
 $ \frac{y_{t}}{l_{t}w_{t}} }  to undo the effect of the other seasonality 
 factor on the observed data value. 
 
 \item \strong{Eq. 3.5. Second Seasonal Factors Adjustment Equation}
 
 The adjustment equation for the second seasonality term is identical to eq. 3.4. 
 with the slight change in the parameters' notations and values. 
 }
 }
 
 @section Prior Distributions for LGT and SGT models:
 
 The default prior distributions of the parameters are given below:
 
 \describe{
 \item{$ \sigma,\gamma,\varsigma }}{Cauchy distribution with 0 location value and the scale parameter equals to 1/200 of the maximum value of y}
 \item{$b}}{Normally distributed with a mean of 0 and standard deviation of 1/200 of the maximum value of y.}
 \item{$\phi}}{Uniformly distributed between -1 and 1}
 \item{$\alpha,\beta,\zeta, \delta}}{Uniform between 0 and 1}
 \item{$ s_{t}, w_{t}}}{ (i.i.d) normal with a mean of 1 and standard deviation of 0.3 before being normalised}
 \item{$\rho}}{Uniform between -0.5 and 1.0}
 \item{$v}}{Uniformly distributed between 2 and 20}
 \item{$\tau}}{Beta distribution with shape parameters  $\alpha =1}  and $\beta =1 }}
 }
 
 Note that some of these prior distributions can be adjusted by the users in the \code{rlgt} function
 
 @references 
 Stan Development Team (2017). RStan: the R interface to Stan. R package version 2.16.2. http://mc-stan.org
 


## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
